---
title: "Respirometry Calculations"
subtitle: "Quadra Island British Columbia 2019"
author: "Nathan Spindel"
output:
  pdf_document: default
  html_document:
    keep_md: yes
editor_options:
  chunk_output_type: console
---
## Workflow:
1) Import data, check for common errors, plot for visual inspection, store as object.  
2) Analyze object to estimate desired rate parameter.  
3) Generate diagnostic plots.  
4) Adjust data based on background respiration.  
5) Convert rate to volume or mass-specific rate.

See Harianto, Carey, and Byrne 2019^[1	Harianto, J., Carey, N. & Byrne, M. respR-An R package for the manipulation and analysis of respirometry data. Methods in Ecology and Evolution 0, doi:10.1111/2041-210x.13162 (2019).] for documentation on **respR** package.
# Load packages and utility functions
```{r include=FALSE}
# Set working directory:
# setwd('C:/Users/nates/OneDrive/Documents/Respirometry/2019_Quadra')
working_directory <- file.path('G:', 'My Drive', 'Quadra')
setwd(working_directory)
# Load packages:
library(tidyverse)
library(readxl)
library(lubridate)
library(respirometry)
library(respR)
library(progress)
library(ggthemes)
# Store file path prefix for convenience.
# prefix <- "C:/Users/nates/OneDrive/Documents/Respirometry/2019_Quadra/Data/"
prefix <- paste0(working_directory, "/5_Data/Presens_Respirometry_Files/")
```

```{r}
calc_metabolic_rate <- function(sample.id, metric) {
  sample.id = i
  # Subset biometric data for just the focal individual
  mass.data <- biometrics %>%
    filter(sample_id == sample.id)
  # Get the label for the focal individual
  sample.label <- mass.data$label
  # Create a string for the filename of the respiration data for the focal individual
  suffix <- paste(sample.label, ".xlsx", sep = "", collapse = NULL)
  respiration.file = paste(prefix, suffix, sep = "", collapse = NULL)
  # Create a string for the filename of the background respiration data associated with the focal individual
  background.row <- biometrics %>%
    filter(
      date == mass.data$date
      &
        species == mass.data$species
      &
        is_background == 'Y'
      &
        site == mass.data$site
      &
        trial_no == mass.data$trial_no
      &
        habitat == mass.data$habitat
      &
        chamber_vol_ml == mass.data$chamber_vol_ml
    )
  background.label <- background.row$label
  suffix.background <- paste(background.label, ".xlsx", sep = "", collapse = NULL)
  background.file = paste(prefix, suffix.background, sep = "", collapse = NULL)
  
  whole.dry <- mass.data$gonad_dry_mass_g + mass.data$animal_dry_mass_g
  # Ash-free dry weight of single gonad in grams.
  single.gonad.afdw <- mass.data$gonad_dry_mass_g - mass.data$gonad_ash_free_dry_wt_g
  # Ash-free dry weight in grams of all 5 gonads combined.
  cumulative.gonad.afdw <- single.gonad.afdw * 5
  # Ash-free dry weight in grams: whole animal dry mass - whole animal afdw + single gonad afdw
  whole.afdw <- mass.data$animal_dry_mass_g - mass.data$animal_ash_free_dry_wt_g - single.gonad.afdw
  # Total volume of chamber in liters
  chamber.volume <- mass.data$chamber_vol_ml / 1000 
  # Calculate volume of individual in liters assuming oblate spheroid geometry.
  if (is.na(mass.data$test_diam_mm_tissue_sampling)) {
    test.diam = mass.data$test_diam_mm
    test.ht = mass.data$test_ht_mm
  } else {
    test.diam = mass.data$test_diam_mm_tissue_sampling
    test.ht = mass.data$test_ht_mm_tissue_sampling
  }
  
  semi.major.axis.mm <- 0.5 * test.diam
  semi.minor.axis.mm <- 0.5 * test.ht
  spheroid.volume.ml <- (4/3 * pi * semi.major.axis.mm ^2 * semi.minor.axis.mm)/1000 # # Equation for volume of oblate spheroid = 4/3pia^2c in mL
  # Store volume of individual based on displacement volume.
  displacement.volume.ml <- mass.data$displacement_vol_ml
  # Get volume of specimen. Choose the measured displacement volume if it exists, otherwise use the volume calculated
  # based on oblate spheroid geometry:
  if (is.na(mass.data$displacement_vol_ml)) {
    specimen.volume <<- spheroid.volume.ml
  } else {
    specimen.volume <<- mass.data$displacement_vol_ml
  }
  # # Ash-free dry weight of single gonad in grams.
  # single.gonad.afdw <- mass.data$gonad_dry_mass_g - mass.data$gonad_ash_free_dry_wt_g
  # # Ash-free dry weight in grams of all 5 gonads combined.
  # cumulative.gonad.afdw <- single.gonad.afdw * 5
  # # Ash-free dry weight in grams: whole animal dry mass - whole animal afdw + single gonad afdw
  # whole.afdw <- mass.data$animal_dry_mass_g - mass.data$animal_ash_free_dry_wt_g + single.gonad.afdw
  # # Total volume of chamber in liters
  # chamber.volume <- mass.data$chamber_vol_ml / 1000 
  # # Calculate volume of individual assuming oblate spheroid geometry.
  # semi.major.axis.mm <- 0.5 * mass.data$test_diam_mm_tissue_sampling
  # semi.minor.axis.mm <- 0.5 * mass.data$test_ht_mm_tissue_sampling
  # spheroid.volume.ml <- (4/3 * pi * semi.major.axis.mm ^2 * semi.minor.axis.mm)/1000 # # Equation for volume of oblate spheroid = 4/3pia^2c in mL
  # # Store volume of individual based on displacement volume.
  # displacement.volume.ml <- mass.data$displacement_vol_ml
  
  # Get background microbial respiration background data:
  background <- read_excel(path = background.file, sheet = "SADD0002000036, Ch 2")
  
  # Import and inspect
  background.inspected <- inspect(background, time = 1, oxygen = 7) # In PreSens default output, the oxygen measurement is in the 7th column
  # Determine optimal linear segment (see respR::auto_rate) 
  background.rate <- auto_rate(background.inspected)
  # Get the start time point for the optimal linear segment of the background signal
  optimal.start <- background.rate$summary$time[1]
  # Get the end time point for the optimal linear segment of the background signal
  optimal.end <- background.rate$summary$endtime[1]
  # Use subset determined using respR::auto_rate to calculate background rate of change in oxygen concentration
  bg <- calc_rate.bg(background, xcol = 1, ycol = 7, from = optimal.start, to = optimal.end, by = "time")
  
  urchin <- read_excel(path = respiration.file, sheet = 6) # sheet 6 contains the oxygen measurement data
  
  # Import and inspect
  urchin.inspected <- inspect(urchin, time = 1, oxygen = 7) # In PreSens default output, the oxygen measurement is in the 7th column
  # Determine optimal linear segment (see respR::auto_rate) 
  urchin.rate <- auto_rate(urchin.inspected)
  
  # Adjust rate
  adj.urchin.rate <- adjust_rate(urchin.rate, bg)
  
  # See respR::convert_rate
  # convert_rate(x, o2.unit = NULL, time.unit = NULL, output.unit = NULL,
  #   volume = NULL, mass = NULL, S = NULL, t = NULL, P = NULL)
  # Note: A string-matching algorithm is used to identify units. 
  # E.g. all of these are the same: mg/L; mg/l, mg L-1, mgL-1, mg per litre, mg.l-1, mg.L-1
  # 
  # O2 Units - Do not require t, S and P
  # [1] "mg/L"   "ug/L"   "mmol/L" "umol/L"
  # 
  # O2 Units - Require t, S and P
  #  [1] "mL/L"    "mg/kg"   "ug/kg"   "mmol/kg" "umol/kg" "mL/kg"   "%"       "Torr"    "hPa"     "kPa"     "mmHg"    "inHg"   
  # 
  # Time units
  # [1] "s" "m" "h"
  # 
  # Output mass units
  # [1] "ug" "mg" "g"  "kg"
  urchin_MO2 <- convert_rate(adj.urchin.rate,
                             o2.unit = "mg/l", # dissolved oxygen unit used to determine rate
                             time.unit = "s", # time unit of the data used to determine rate
                             output.unit = "mg/m/kg", 
                             volume = chamber.volume - specimen.volume, # total volume of chamber in liters excluding the specimen volume
                             mass = metric) # mass of sample in kg. (e.g. ash-free dry weight, dry weight, etc).
  # Store metabolic rate data for focal individual in data frame.
  sample.result <- tibble(
    sample_id = mass.data$sample_id, 
    date = mass.data$date,
    label = mass.data$label,
    species = mass.data$species,
    habitat = mass.data$habitat,
    site = mass.data$site,
    displacement_volume_ml = mass.data$displacement_vol_ml,
    spheroid_volume_ml = spheroid.volume.ml,
    whole_ash_free_dry_wt_g = whole.afdw,
    cumulative_gonad_ash_free_dry_wt_g = cumulative.gonad.afdw,
    single_gonad_ash_free_dry_wt_g = single.gonad.afdw,
    whole_dry_wt_g = whole.dry,
    cumulative_gonad_dry_wt_g = mass.data$gonad_dry_mass_g * 5,
    single_gonad_dry_wt_g = mass.data$gonad_dry_mass_g,
    input_rate = urchin_MO2$summary$input.rate[1],
    converted_rate = urchin_MO2$summary$converted.rate[1],
    volumetric_rate = urchin_MO2$summary$volumetric[1],
    mass_specific_rate = urchin_MO2$summary$mass.specific[1],
    rate_unit = urchin_MO2$output.unit,
    respiration_filename = respiration.file
  )
  
  if (!is_null(sample.result)) {
    return(sample.result)
  } else {
    print(paste0("No result calculated for sample id: ", sample.id))
  }
}
```


# Get general biometric data.
```{r}
# Get full set of biometric data.
biometrics <- read_excel(
  # path = "C:/Users/nates/OneDrive/Documents/Respirometry/2019_Quadra/Data/Urchin_Biometrics/Urchin_Biometrics.xlsx", 
  path = paste(working_directory, "2019_Haida_Gwaii/Data/Urchin_Biometrics/Urchin_Biometrics.xlsx", sep = "/"),
  sheet = "data")
```
# Unique identifier for the focal individual.
```{r}
sample.id = 3
```
# Gather quirks:
```{r}
# Store sample.id's that are problematic for later troubleshooting
# Whole animal ash-free dry weight quirks:
quirks = c()
```
# Get sample id's for non-background samples.
```{r}
# Get the sample id's which are not background samples and not included in the list of quirks.
# Probable cause: chosen normalization metric not available for focal individual (i.e. must select alternative
# normalization metric).
not.background <- biometrics %>%
  filter(is_background == "N" & !sample_id %in% quirks) %>%
  select(sample_id)
```
# Get biometric data for focal individual.
```{r collapse=TRUE, tidy=TRUE}
# Subset biometric data for just the focal individual
mass.data <- biometrics %>%
 filter(sample_id == sample.id)
# Get the label for the focal individual
sample.label <- mass.data$label
# Create a string for the filename of the respiration data for the focal individual
suffix <- paste(sample.label, ".xlsx", sep = "", collapse = NULL)
respiration.file = paste(prefix, suffix, sep = "", collapse = NULL)
# Create a string for the filename of the background respiration data associated with the focal individual
background.row <- biometrics %>%
  filter(
    date == mass.data$date
    &
    species == mass.data$species
    &
    is_background == 'Y'
    &
    site == mass.data$site
    &
    trial_no == mass.data$trial_no
    &
    habitat == mass.data$habitat
    &
    chamber_vol_ml == mass.data$chamber_vol_ml
  )
background.label <- background.row$label
suffix.background <- paste(background.label, ".xlsx", sep = "", collapse = NULL)
background.file = paste(prefix, suffix.background, sep = "", collapse = NULL)
```
Get biometric data to be used for normalizing oxygen consumption rate.
```{r collapse=TRUE, tidy=TRUE}
whole.dry <- mass.data$gonad_dry_mass_g + mass.data$animal_dry_mass_g
# Ash-free dry weight of single gonad in grams.
single.gonad.afdw <- mass.data$gonad_dry_mass_g - mass.data$gonad_ash_free_dry_wt_g
# Ash-free dry weight in grams of all 5 gonads combined.
cumulative.gonad.afdw <- single.gonad.afdw * 5
# Ash-free dry weight in grams: whole animal dry mass - whole animal afdw + single gonad afdw
whole.afdw <- mass.data$animal_dry_mass_g - mass.data$animal_ash_free_dry_wt_g - single.gonad.afdw
# Total volume of chamber in liters
chamber.volume <- mass.data$chamber_vol_ml / 1000 
# Calculate volume of individual in liters assuming oblate spheroid geometry.
if (is.na(mass.data$test_diam_mm_tissue_sampling)) {
  test.diam = mass.data$test_diam_mm
  test.ht = mass.data$test_ht_mm
} else {
  test.diam = mass.data$test_diam_mm_tissue_sampling
  test.ht = mass.data$test_ht_mm_tissue_sampling
}

semi.major.axis.mm <- 0.5 * test.diam
semi.minor.axis.mm <- 0.5 * test.ht
spheroid.volume.ml <- (4/3 * pi * semi.major.axis.mm ^2 * semi.minor.axis.mm)/1000 # # Equation for volume of oblate spheroid = 4/3pia^2c in mL
# Store volume of individual based on displacement volume.
displacement.volume.ml <- mass.data$displacement_vol_ml
# Get volume of specimen. Choose the measured displacement volume if it exists, otherwise use the volume calculated
# based on oblate spheroid geometry:
if (is.na(mass.data$displacement_vol_ml)) {
  specimen.volume <<- spheroid.volume.ml
} else {
  specimen.volume <<- mass.data$displacement_vol_ml
}
```
Get background data.
```{r collapse=TRUE}
# Get background microbial respiration background data:
# background <- read_excel(path = background.file, sheet = "SADD0002000036, Ch 2")
background <- read_excel(path = background.file, sheet = "SADD0002000036, Ch 1")
```
Select linear segment of background data, calculate rate of change in oxygen concentration.
```{r}
# Import and inspect
background.inspected <- inspect(background, time = 1, oxygen = 7) # In PreSens default output, the oxygen measurement is in the 7th column
# Determine optimal linear segment (see respR::auto_rate) 
background.rate <- auto_rate(background.inspected)
# Get the start time point for the optimal linear segment of the background signal
optimal.start <- background.rate$summary$time[1]
# Get the end time point for the optimal linear segment of the background signal
optimal.end <- background.rate$summary$endtime[1]
# Use subset determined using respR::auto_rate to calculate background rate of change in oxygen concentration
bg <- calc_rate.bg(background, xcol = 1, ycol = 7, from = optimal.start, to = optimal.end, by = "time")
```
Get focal organism respiration data.
```{r}
urchin <- read_excel(path = respiration.file, sheet = 6) # sheet 6 contains the oxygen measurement data
```
Select linear segment of focal organism data, calculate rate of change in oxygen concentration.
```{r}
# Import and inspect
urchin.inspected <- inspect(urchin, time = 1, oxygen = 7) # In PreSens default output, the oxygen measurement is in the 7th column
# Determine optimal linear segment (see respR::auto_rate) 
urchin.rate <- auto_rate(urchin.inspected)
```
Adjust rate calculation to account for background change in oxygen concentration.
```{r}
# Adjust rate
adj.urchin.rate <- adjust_rate(urchin.rate, bg)
```
# Set Metric (e.g. ash-free dry weight, wet weight, etc) for mass normalization of respiration rate.
```{r}
# metric <<- single.gonad.afdw
# metric <<- cumulative.gonad.afdw
metric <<- whole.afdw
# metric <<- whole.dry
```
# Iterate over batch of respiration files, calculate rates, append results to data frame:
```{r}
pb <- progress_bar$new(total = max(not.background$sample_id))

for (i in not.background$sample_id) {
  pb$tick()
  if (!exists("metabolic.data"))
  {
    metabolic.data <- calc_metabolic_rate(i, metric)
  } else {
    metabolic.data <- rbind(metabolic.data, calc_metabolic_rate(i, metric))
  }
}
```
# Summarize and visualize results:
```{r}
# Aggregate mass-specific and volumetric rate by species, and habitat.
# Calculate summary statistics.
summary.metabolic.rate <- metabolic.data %>%
  group_by(species, habitat) %>%
  summarise(
    n = n(),
    avg_mass_specific_rate = mean(mass_specific_rate, na.rm = TRUE),
    sd_mass_specific_rate = sd(mass_specific_rate, na.rm = TRUE),
    avg_volumetric_rate = mean(volumetric_rate, na.rm = TRUE),
    sd_volumetric_rate = sd(volumetric_rate, na.rm = TRUE)
  ) %>%
  mutate(
    se_mass_specific_rate = sd_mass_specific_rate/sqrt(n),
    se_volumetric_rate = sd_volumetric_rate/sqrt(n)
  )

summary.metabolic.rate

# Visualize distribution of rate data.
# Histogram mass-specific rate x species x habitat x site:
p1 <- ggplot(metabolic.data, aes(mass_specific_rate)) +
        geom_histogram(fill = "white", color = "grey30", bins = 25) +
        facet_wrap(vars(species, habitat, site))

p1
# Boxplot mass-specific rate x species x habitat x site:
p2 <- ggplot(metabolic.data) +
        geom_boxplot(aes(x = habitat, y = mass_specific_rate, fill = habitat), color = "grey30") +
        facet_wrap(vars(species, site)) +
        xlab("") +
        ylab("Oxygen consumption rate (mg O2/min/kg ash-free dry tissue mass)") +
        ggtitle("Mass-Specific Metabolic Rate (MO2)") +
        theme_pander()

p2
# Histogram volumetric rate x species x habitat x site:
p3 <- ggplot(metabolic.data, aes(volumetric_rate)) +
        geom_histogram(fill = "white", color = "grey30", bins = 25) +
        facet_wrap(vars(species, habitat, site)) +
        ggtitle("Volumetric Metabolic Rate (VO2)") +
        theme_pander()

p3
# Boxplot volumetric rate x species x habitat x site:
p4 <- ggplot(metabolic.data) +
        geom_boxplot(aes(x = habitat, y = volumetric_rate, fill = habitat), color = "grey30") +
        facet_wrap(vars(species, site)) +
        ggtitle("Volumetric Metabolic Rate (VO2)") +
        theme_pander()

p4
# Get red data
red.metab <- metabolic.data %>%
  filter(species == "Mesocentrotus franciscanus")

# Get green data
green.metab <- metabolic.data %>%
  filter(species == "Strongylocentrotus droebachiensis")

# Scatterplot oxygen consumption rate x volume x species x habitat:
p5 <- ggplot(red.metab) +
  geom_point(aes(x = displacement_volume_ml, y = 0-converted_rate)) +
  # geom_point(aes(x = displacement_volume_ml, y = 0-converted_rate)) +
  # geom_point(aes(x = spheroid_volume_ml, y = 0-converted_rate)) +
  # geom_smooth(aes(x = spheroid_volume_ml, y = 0-converted_rate)) +
  facet_wrap(vars(habitat)) +
  xlab("Ash-free Dry Mass (mg)")
  # xlab("Displacement volume (mL)") +
  # xlab("Oblate spheroid test volume (mL)") +
  ylab("Oxygen consumption rate mg O2/min/kg tissue") +
  ggtitle("Mesocentrotus franciscanus") +
  theme_tufte()

p5

p6 <- ggplot(green.metab) +
  geom_point(aes(x = displacement_volume_ml, y = 0-converted_rate)) +
  # geom_point(aes(x = spheroid_volume_ml, y = 0-converted_rate)) +
  # geom_smooth(aes(x = spheroid_volume_ml, y = 0-converted_rate)) +
  facet_wrap(vars(habitat)) +
  xlab("Displacement volume (mL)") +
  # xlab("Oblate spheroid test volume (mL)") +
  ylab("Oxygen consumption rate mg O2/min/kg tissue") +
  ggtitle("Strongylocentrotus droebachiensis") +
  theme_tufte()

p6

# Visualize summary results.
# Normalized by ash-free dry weight:
summary.mass.specific.plot <- ggplot(data = summary.metabolic.rate) +
  geom_bar(aes(x = habitat, y = avg_mass_specific_rate, fill = habitat), stat = "identity") +
  facet_wrap(vars(species)) +
  geom_errorbar(aes(x = habitat, ymin = avg_mass_specific_rate - se_mass_specific_rate, ymax = avg_mass_specific_rate + se_mass_specific_rate), width = 0.2) +
  ggtitle("Ash-Free Dry Mass-Normalized Metabolic Rate (MO2)") +
  ylab("Oxygen consumption rate (mg O2/min/kg tissue)") +
  xlab("") +
  theme_minimal()

summary.mass.specific.plot

# Volumetric:
summary.volumetric.plot <- ggplot(data = summary.metabolic.rate) +
  geom_bar(aes(x = habitat, y = avg_volumetric_rate, fill = habitat), stat = "identity") +
  facet_wrap(vars(species)) +
  geom_errorbar(aes(x = habitat, ymin = avg_volumetric_rate - se_volumetric_rate, ymax = avg_volumetric_rate + se_volumetric_rate), width = 0.2) +
  ggtitle("Volumetric Metabolic Rate (VO2)") +
  ylab("Oxygen consumption rate (mg O2/min/kg tissue)") +
  xlab("") +
  theme_minimal()

summary.volumetric.plot
```
# Write data to xlsx file:
```{r}
library(writexl)
# Whole animal ash-free dry weight-normalized metabolic rate data:
write_xlsx(metabolic.data, path = "./Data/Calculations/whole_afdw_metabolic_rate.xlsx")
```